Identifying Effective Motivational Interviewing Sequences Using Automated Pattern Analysis

### 2018-04-14 jhir/latex-manuscript/manuscript.tex
update manuscript

\begin{itemize}
\item $M$ is the number of distinct observations symbols per state, i.e. the discrete codebook size (Table~\ref{tab:codebook})
\item $T$ is a state transition probability matrix, in which is the probability of HMM transitioning from state to state  
\item $E$ is an emission probability matrix, in which is the probability of observing symbol, when HMM is in state 
\item $p$ is the initial state distribution vector where is the probability of the first HMM state
\end{itemize}

@@

\section{Results}
\label{sec:results}
The transition and emission probability matrices of HMM model are reported in Tables~\ref{tab:emission} and \ref{tab:transition}. The three behaviors with the highest frequency, representing 45-69\% of each state's emission probability mass, were used to interpret the emission matrix and label the hidden states. Most (84\%) successful sequences began with a state characterized as \textit{``high motivation''} as evidenced by a greater proportion of three counselor behaviors: reflections of change talk (29\%), reflections of commitment language (17\%), and affirmations (12\%). Successful sequences began with a state of \textit{``high receptivity''} 11\% of the time. \textit{``High receptivity''} sequences were characterized by nearly equal proportions of information offered using patient-centered strategies (18\%), questions to elicit change talk (16\%), and affirmations (15\%). Few successful sequences began in states of \textit{``moderate receptivity''} and \textit{``low receptivity''} (2\% and 3\% of the time, respectively). These two states were characterized by different proportions of the same behaviors. \textit{Moderate receptivity} sequences were distinguished from \textit{low receptivity} sequences by a greater proportion of counselor questions to elicit change talk (20\% versus 11\%) and a lower proportion of patient low uptake statements (16\% versus 28\%); counselor statements emphasizing the patient's autonomy were about the same (14\% versus 17\%). No (0\%) successful sequence began in the \textit{``active feedback''} state, which was characterized by three patient behaviors, low uptake (47\%), weight-related high uptake (12\%), and other-related high uptake (10\%). Successful sequences transitioned from \textit{``high motivation''} to \textit{``active feedback''} most often (41\%). \textit{``active feedback''}, in turn, most frequently transitioned to \textit{``high receptivity''} (39\%). The \textit{``moderate receptivity''} state transitioned to most often to \textit{``active feedback''} (27\%) and automatically transitioned back to \textit{``moderate receptivity''} (24\%) or to \textit{``low receptivity''} (23\%) with similar frequency. The full transition matrix is presented in Table~\ref{tab:transition}.

@@

Frequent communication patterns in successful and unsuccessful patient-counselor communication sequences are shown in Table~\ref{tab:patterns}. Results from frequent pattern mining analysis indicate that reflections of change talk were the most frequent counselor communication behavior in both successful (36.1\%) and unsuccessful sequences (33.7\%). For each pattern, the statistical significance of difference between pattern occurrence in successful and unsuccessful communications was reported by computing p-value using Pearson's chi-square test. Successful sequences were distinguished from unsuccessful sequences by a higher frequency of counselor questions phrased to elicit change talk (30.8\% versus 17.4\%, p= $<$0.0001), statements emphasizing the patient's decision-making autonomy (28.5\% versus 18.6\%, p=0.0012), questions phrased to elicit commitment language (18.1\% versus 11.6\%, p=0.0118), and reflections of commitment language (20.7\% versus 15.1\%, p= 0.0423). In contrast, unsuccessful sequences were characterized by greater frequency of questions to elicit perceived barriers (14.7\% versus 0\%, p= $<$0.0001), reflections of sustain talk (27.1\% versus 15.8\%, p= $<$0.0001), providing information (28.7\% versus 22.1\%, p= 0.0254) and other reflections (11.6\% versus 0\%, p= $<$0.0001). In 14.0\% of the successful sequences, reflections of change talk were paired with a question phrased to elicit change talk; this pattern did not appear in unsuccessful sequences. In contrast, in 10.5\% of the unsuccessful sequences, reflections of change talk were paired with information; this pattern did not appear in successful sequences.

>>>>

\begin{itemize}
\item $M$ is the number of distinct observations symbols per state, i.e. the discrete codebook size (Table~\ref{tab:codebook})
\item $T$ is an $NXN$ state transition probability matrix, in which $t_{ij}$ is the probability of HMM transitioning from state $i$ to state $j$  
\item $E$ is an $NXM$ emission probability matrix, in which $e_{jk}$ is the probability of observing symbol $k$, when HMM is in state $j$ 
\item $p$ is the initial state distribution vector where $\pi_i$ is the probability of the first HMM state
\end{itemize}

@@

\section{Results}
\label{sec:results}
The transition and emission probability matrices of HMM model are reported in Tables~\ref{tab:emission} and \ref{tab:transition}. The three behaviors with the highest frequency, representing 45-69\% of each state's emission probability mass, were used to interpret the emission matrix and label the hidden states. Most (84\%) successful sequences began with a state characterized as \textit{``high motivation''} as evidenced by a greater proportion of three counselor behaviors: reflections of change talk (29\%), reflections of commitment language (17\%), and affirmations (12\%). Successful sequences began with a state of \textit{``high receptivity''} 11\% of the time. \textit{``High receptivity''} sequences were characterized by nearly equal proportions of information offered using patient-centered strategies (18\%), questions to elicit change talk (16\%), and affirmations (15\%). Few successful sequences began in states of \textit{``moderate receptivity''} and \textit{``low receptivity''} (2\% and 3\% of the time, respectively). These two states were characterized by different proportions of the same behaviors. \textit{``Moderate receptivity''} sequences were distinguished from \textit{``low receptivity''} sequences by a greater proportion of counselor questions to elicit change talk (20\% versus 11\%) and a lower proportion of patient low uptake statements (16\% versus 28\%); counselor statements emphasizing the patient's autonomy were about the same (14\% versus 17\%). No (0\%) successful sequence began in the \textit{``active feedback''} state, which was characterized by three patient behaviors, low uptake (47\%), weight-related high uptake (12\%), and other-related high uptake (10\%). Successful sequences transitioned from \textit{``high motivation''} to \textit{``active feedback''} most often (41\%). \textit{``active feedback''}, in turn, most frequently transitioned to \textit{``high receptivity''} (39\%). The \textit{``moderate receptivity''} state transitioned to most often to \textit{``active feedback''} (27\%) and automatically transitioned back to \textit{``moderate receptivity''} (24\%) or to \textit{``low receptivity''} (23\%) with similar frequency. The full transition matrix is presented in Table~\ref{tab:transition}.

@@

Frequent communication patterns in successful and unsuccessful patient-counselor communication sequences are shown in Table~\ref{tab:patterns}. Results from frequent pattern mining analysis indicate that reflections of change talk were the most frequent counselor communication behavior in both successful (36.1\%) and unsuccessful sequences (33.7\%). For each pattern, the statistical significance of difference between pattern occurrence in successful and unsuccessful communications was reported by computing p-value using Pearson's chi-square test. Successful sequences were distinguished from unsuccessful sequences by a higher frequency of counselor questions phrased to elicit change talk (30.8\% versus 17.4\%, p= $<$0.0001), statements emphasizing the patient's decision-making autonomy (28.5\% versus 18.6\%, p=0.0012), questions phrased to elicit commitment language (18.1\% versus 11.6\%, p=0.0118), and reflections of commitment language (20.7\% versus 15.1\%, p= 0.0423). In contrast, unsuccessful sequences were characterized by greater frequency of questions to elicit perceived barriers (14.7\% versus 0\%, p= $<$0.0001), reflections of sustain talk (27.1\% versus 15.8\%, p= $<$0.0001), providing information (28.7\% versus 22.1\%, p=0.0254) and other reflections (11.6\% versus 0\%, p= $<$0.0001). In 14.0\% of the successful sequences, reflections of change talk were paired with a question phrased to elicit change talk; this pattern did not appear in unsuccessful sequences. In contrast, in 10.5\% of the unsuccessful sequences, reflections of change talk were paired with information; this pattern did not appear in successful sequences.
